Implications of taxonomic bias for microbial differential-abundance analysis

### 2022-08-19 v/3bde34f650d7c06e05bd5957ebca2a7857a4c0ed/_main.tex
Make edits from AW and BC
This commit adds the rendered manuscript associated with
3bde34f650d7c06e05bd5957ebca2a7857a4c0ed

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis},
  pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

@@

% Macros
\newcommand{\md}{^{\times}\!/}
\newcommand{\tot}{{\text{tot}}}
\newcommand{\mtot}{{\text{[tot]}}}
\DeclareMathOperator{\FE}{FE}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis}
\author{Michael R. McLaren\footnote{Massachusetts Institute of Technology; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T. Nearing\footnote{Dalhousie University} \and Amy D. Willis\footnote{University of Washington} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2022-07-17}

@@

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{preface}{%
\section*{Preface}\label{preface}}
\addcontentsline{toc}{section}{Preface}

\begin{verbatim}
#> working directory clean
\end{verbatim}

\emph{This manuscript was rendered from commit 9c86fbcc416ebe9d18d2e80120064bea4c346f36.}

\leavevmode\vadjust pre{\hypertarget{preface-warning}{}}%
\textbf{This in-progress manuscript is not intended for general scientific use.}
It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
Please post comments or questions on the \href{https://github.com/mikemc/differential-abundance-theory/issues}{GitHub Issues page} or \href{m.mclaren42@gmail.com}{email Mike}.

This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
Analyses of both relative and absolute abundances are considered.
The manuscript is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.
Supporting data analyses can be found in the accompanying \href{https://mikemc.github.io/differential-abundance-theory/notebook/}{computational research notebook}.

@@

\leavevmode\vadjust pre{\hypertarget{abstract}{}}%
\textbf{Abstract:}

@@

The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.
If bias caused a species' abundance to be consistently measured as 10-fold greater than its actual value, then we would still recover the correct FDs among samples.
However, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed mathematically and with MGS measurements of artificially constructed (`mock') communities that consistent taxonomic bias can create fold errors (FEs) that vary across samples and, as a result, majorly distort cross-sample comparisons.
In particular, they showed that the FE in a species' proportion---the most common measure of relative abundance---varies among samples, distorting the observed FDs between samples.
In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods (for relative and absolute abundance) that are based on species proportions.
We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
We further demonstrate that another set of DA methods, based on the ratios among species, are robust to bias.
Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
Application of these insights and methods can build the confidence needed to turn the findings of microbiome studies into readily-translatable scientific knowledge.

@@

This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe the effect that consistent taxonomic bias within an MGS experiment affects the relative and absolute abundances measured for various microbial species.
We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.

\hypertarget{model-of-mgs-measurement}{%
\subsection{Model of MGS measurement}\label{model-of-mgs-measurement}}

@@

We distinguish between two types of species-level \emph{relative abundances} within a sample.
The \emph{proportion} \(P_{i}^{(a)}\) of species \(i\) in sample \(a\) equals its abundance relative to the total abundance of all species in \(S\),
\begin{align}
  \label{eq:prop}
  P_{i}^{(a)} &\equiv \frac{A_i^{(a)}}{A_{\text{tot}}^{(a)}}.
\end{align}
The \emph{ratio} \(R_{i/j}^{(a)}\) between two species \(i\) and \(j\) equals the abundance of \(i\) divided by that of \(j\),
\begin{align}
  \label{eq:ratio}
  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
\end{align}

The measured proportion of a species is given by its proportion of all the assigned reads in a sample,
\begin{align}
  \label{eq:prop-meas}
  \tilde P_{i}^{(a)} &\equiv \frac{M_i^{(a)}}{M_{\text{tot}}^{(a)}}.
\end{align}
We use the tilde to distinguish the measurement from the actual quantity being measured.
From Equations \eqref{eq:mgs-model}, \eqref{eq:total-reads}, and \eqref{eq:prop-meas}, it follows that the measured and actual proportion are related by
\begin{align}
  \label{eq:prop-error}
  \tilde P_{i}^{(a)} &= P_{i}^{(a)} \cdot \frac{B_i}{\bar B^{(a)}}.
\end{align}
Taxonomic bias thus creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
Since the mean efficiency varies across samples, so does the FE.
This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \ref{fig:error-proportions}.
Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE \textless{} 1) but over-measured (FE \textgreater{} 1) in Sample 2.
This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3C} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).

The measured ratio \(\tilde R_{i/j}^{(a)}\) between species \(i\) and \(j\) is given by the ratio of their read counts,
\begin{align}
  \label{eq:ratio-meas}
  \tilde R_{i/j}^{(a)} \equiv \frac{M_{i}^{(a)}}{M_{j}^{(a)}}.
\end{align}
From Equations \eqref{eq:mgs-model} and \eqref{eq:ratio-meas}, it follows that the measured and actual ratio are related by
\begin{align}
  \label{eq:ratio-error}
  \tilde R_{i/j}^{(a)} = R_{i/j}^{(a)} \cdot \frac{B_i}{B_j}.
\end{align}
Taxonomic bias thus creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3D} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates fold errors in species proportions that vary across samples and lead to inaccurate fold differences between samples.} Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold difference in the third species that is derived from these measurements. Species' proportions may be measured as too high or too low depending on sample composition. For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).}\label{fig:error-proportions}
\end{figure}

\textbf{Higher-order taxa:}
We can consider a higher-order taxon \(I\), such as a genus or phylum, as a set of species, \(\{i \in I\}\).
The abundance of taxon \(I\) in sample \(a\) is the sum of the abundances of its constituent species, \(A_{I}^{(a)} \equiv \sum_{i \in I} A_{i}^{(a)}\).
Similarly, the read count of taxon \(I\) is the sum \(M_{I}^{(a)} \equiv \sum_{i \in I} M_{i}^{(a)}\).
We further define the efficiency of taxon \(I\) as the abundance-weighted average of the efficiencies of its constituent species,
\begin{align}
  \label{eq:efficiency-general}
  B_I^{(a)} \equiv \frac{\sum_{i\in I} A_{i}^{(a)} B_{i}}{\sum_{i\in I} A_{i}^{(a)}}.
\end{align}
With these definitions, the read count for taxon \(I\) can be expressed as
\(M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}\).
Thus \(B_I^{(a)}\) plays a role analogous to the efficiency of an individual species, but differs in that it need not be constant across samples:
If the constituent species have different efficiencies, then the efficiency of the higher-order taxon \(I\) depends on the relative abundances of its constituents and so will tend to vary across samples (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
As an example, suppose that Species 1 and Species 2 in Figure \ref{fig:error-proportions} were in the same phylum.
The efficiency of the phylum would then be \(\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5\) in Sample 1 and \(\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1\) in Sample 2.
Equations \eqref{eq:prop-error} and \eqref{eq:ratio-error} continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent, higher-order taxa efficiencies \(B_I^{(a)}\) and \(B_J^{(a)}\) are used.
In this way, we see that both proportions and ratios among higher-order taxa may have inconsistent FEs.

@@

Several extensions of the standard MGS experiment make it possible to measure absolute species abundances.
These extensions fall into two general approaches.
The first approach leverages information about the abundance of the total community; for example, Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017}) measured total-community abundance using flow cytometry and multiplied this number by MGS genus proportions to obtain the absolute abundances of individual genera (Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017})).
A second approach leverages information about the abundance of one or more individual species; for example, a researcher might `spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (Harrison et al. (\protect\hyperlink{ref-harrison2021theq}{2021})).
We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.

@@

Suppose that the total abundance of all species in the sample, \(A_{{\text{tot}}}^{(a)}\), has been measured by a non-MGS method, yielding a measurement \(\tilde A_{\text{tot}}^{(a)}\).
The absolute abundance of an individual species can be measured by multiplying the species' proportion from MGS by this total-abundance measurement,
\begin{align}
  \label{eq:density-prop-meas}
  \tilde A_i^{(a)} &= \tilde P_i^{(a)} \tilde A_{\text{tot}}^{(a)}.
\end{align}
Total-abundance measurements recently used for this purpose include counting cells with microscopy (Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020})) or flow cytometry (Props et al. (\protect\hyperlink{ref-props2017abso}{2017}), Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020})), measuring the concentration of a marker-gene with qPCR or ddPCR (Zhang et al. (\protect\hyperlink{ref-zhang2017soil}{2017}), Barlow, Bogatyrev, and Ismagilov (\protect\hyperlink{ref-barlow2020aqau}{2020}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020}), Tettamanti Boshier et al. (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020})), and measuring bulk DNA concentration with a florescence-based DNA quantification method (Contijoch et al. (\protect\hyperlink{ref-contijoch2019gutm}{2019})).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

@@

% Macros
\newcommand{\md}{^{\times}\!/}
\newcommand{\tot}{{\text{tot}}}
\newcommand{\mtot}{{\text{[tot]}}}
\DeclareMathOperator{\FE}{FE}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Implications of taxonomic bias for microbial differential-abundance analysis},
  pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Implications of taxonomic bias for microbial differential-abundance analysis}
\author{Michael R. McLaren\footnote{Massachusetts Institute of Technology; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T. Nearing\footnote{Dalhousie University} \and Amy D. Willis\footnote{University of Washington} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2022-08-18}

@@

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

\leavevmode\vadjust pre{\hypertarget{preface-note}{}}%
This manuscript was rendered from commit 3bde34f of \url{https://github.com/mikemc/differential-abundance-theory}.
Supporting data analyses can be found in the accompanying \href{https://mikemc.github.io/differential-abundance-theory/notebook/}{computational research notebook}.
Please post comments or questions on \href{https://github.com/mikemc/differential-abundance-theory}{GitHub}.
The manuscript is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
See the \href{https://github.com/mikemc/differential-abundance-theory/releases}{GitHub Releases} or \href{https://doi.org/10.5281/zenodo.4552717}{Zenodo record} for earlier versions.

@@

@@

The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.
If bias caused a species' abundance to be consistently measured as 10-fold greater than its actual value, then we would still recover the correct FDs among samples.
However, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed mathematically and with MGS measurements of artificially constructed (`mock') communities that consistent taxonomic bias can create fold errors (FEs) that vary across samples and, as a result, majorly distort cross-sample comparisons.
In particular, they showed that the FE in a species' proportion---the most common measure of relative abundance---varies among samples, distorting the observed FDs between samples.
In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that other abundance measures---those based on the ratios among species---have constant FEs and may lead to more robust DA analyses.
The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods that are based on species proportions, even if bias is the same across samples.
We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
We further demonstrate that another set of DA methods, based on the ratios among species, are robust to consistent bias.
Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
These insights and methods will aid microbiome data analysts in turning the findings of microbiome studies into readily-translatable scientific knowledge.

@@

This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe how taxonomic bias in an MGS experiment affects the relative and absolute abundances measured for various microbial species.
We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.

\hypertarget{a-model-of-mgs-measurements}{%
\subsection{A model of MGS measurements}\label{a-model-of-mgs-measurements}}

@@

We distinguish between two types of species-level \emph{relative abundances} within a sample.
The \emph{proportion} \(P_{i}^{(a)}\) of species \(i\) in sample \(a\) equals its abundance divided by the total abundance of all species in \(S\),
\begin{align}
  \label{eq:prop}
  P_{i}^{(a)} &\equiv \frac{A_i^{(a)}}{A_{\text{tot}}^{(a)}}.
\end{align}
The \emph{ratio} \(R_{i/j}^{(a)}\) between two species \(i\) and \(j\) equals the abundance of \(i\) divided by that of \(j\),
\begin{align}
  \label{eq:ratio}
  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
\end{align}

The measured proportion of a species is given by its proportion of all the assigned reads in a sample,
\begin{align}
  \label{eq:prop-meas}
  \tilde P_{i}^{(a)} &\equiv \frac{M_i^{(a)}}{M_{\text{tot}}^{(a)}}.
\end{align}
We use the tilde to distinguish the measurement from the actual quantity being measured.
From Equations \eqref{eq:mgs-model}, \eqref{eq:total-reads}, and \eqref{eq:prop-meas}, it follows that the measured and actual proportion are related by
\begin{align}
  \label{eq:prop-error}
  \tilde P_{i}^{(a)} &= P_{i}^{(a)} \cdot \frac{B_i}{\bar B^{(a)}}.
\end{align}
Taxonomic bias creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
Since the mean efficiency varies across samples, so does the FE.
This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \ref{fig:error-proportions}.
Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE \textless{} 1) but over-measured (FE \textgreater{} 1) in Sample 2.
This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3C} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).

The measured ratio \(\tilde R_{i/j}^{(a)}\) between species \(i\) and \(j\) is given by the ratio of their read counts,
\begin{align}
  \label{eq:ratio-meas}
  \tilde R_{i/j}^{(a)} \equiv \frac{M_{i}^{(a)}}{M_{j}^{(a)}}.
\end{align}
From Equations \eqref{eq:mgs-model} and \eqref{eq:ratio-meas}, it follows that the measured and actual ratio are related by
\begin{align}
  \label{eq:ratio-error}
  \tilde R_{i/j}^{(a)} = R_{i/j}^{(a)} \cdot \frac{B_i}{B_j}.
\end{align}
Taxonomic bias creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-measured by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3D} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).

\begin{figure}
\includegraphics[width=1\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates fold errors in species proportions that vary across samples and lead to inaccurate fold differences between samples.} Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold difference in the third species that is derived from these measurements. Species' proportions may be measured as too high or too low depending on sample composition. For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).}\label{fig:error-proportions}
\end{figure}

\textbf{Higher-order taxa:}
We can consider a higher-order taxon \(I\), such as a genus or phylum, as a set of species, \(\{i \in I\}\).
The abundance of taxon \(I\) in sample \(a\) is the sum of the abundances of its constituent species, \(A_{I}^{(a)} \equiv \sum_{i \in I} A_{i}^{(a)}\).
Similarly, the read count of taxon \(I\) is the sum \(M_{I}^{(a)} \equiv \sum_{i \in I} M_{i}^{(a)}\).
We further define the efficiency of taxon \(I\) as the abundance-weighted average of the efficiencies of its constituent species,
\begin{align}
  \label{eq:efficiency-general}
  B_I^{(a)} \equiv \frac{\sum_{i\in I} A_{i}^{(a)} B_{i}}{\sum_{i\in I} A_{i}^{(a)}}.
\end{align}
With these definitions, the read count for higher-order taxon \(I\) can be expressed as
\(M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}\).
Thus \(B_I^{(a)}\) plays a role analogous to the efficiency of an individual species, but differs in that it is not constant across samples:
If the constituent species have different efficiencies, then the efficiency of the higher-order taxon \(I\) depends on the relative abundances of its constituents and so will vary across samples (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
As an example, suppose that Species 1 and Species 2 in Figure \ref{fig:error-proportions} were in the same phylum.
The efficiency of the phylum would then be \(\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5\) in Sample 1 and \(\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1\) in Sample 2.
Equations \eqref{eq:prop-error} and \eqref{eq:ratio-error} continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent, higher-order taxa efficiencies \(B_I^{(a)}\) and \(B_J^{(a)}\) are used.
In this way, we see that both proportions and ratios among higher-order taxa may have inconsistent FEs.

@@

Several extensions of the standard MGS experiment make it possible to measure absolute species abundances.
These extensions fall into two general approaches.
The first approach leverages information about the abundance of the total community; for example, Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017}) measured total-community abundance using flow cytometry and multiplied this number by genus proportions measured by MGS to quantify the absolute abundances of individual genera (Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017})).
A second approach leverages information about the abundance of one or more individual species; for example, a researcher might `spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (Harrison et al. (\protect\hyperlink{ref-harrison2021theq}{2021})).
We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.

@@

Suppose that the total abundance of all species in the sample, \(A_{{\text{tot}}}^{(a)}\), has been measured by a non-MGS method, yielding a measurement \(\tilde A_{\text{tot}}^{(a)}\).
The absolute abundance of an individual species can be quantified by multiplying the species' proportion from MGS by this total-abundance measurement,
\begin{align}
  \label{eq:density-prop-meas}
  \tilde A_i^{(a)} &= \tilde P_i^{(a)} \tilde A_{\text{tot}}^{(a)}.
\end{align}
Total-abundance measurements recently used for this purpose include counting cells with microscopy (Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020})) or flow cytometry (Props et al. (\protect\hyperlink{ref-props2017abso}{2017}), Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020})), measuring the concentration of a marker-gene with qPCR or ddPCR (Zhang et al. (\protect\hyperlink{ref-zhang2017soil}{2017}), Barlow, Bogatyrev, and Ismagilov (\protect\hyperlink{ref-barlow2020aqau}{2020}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020}), Tettamanti Boshier et al. (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020})), and measuring bulk DNA concentration with a florescence-based DNA quantification method (Contijoch et al. (\protect\hyperlink{ref-contijoch2019gutm}{2019})).
